Equalizer Programming Guide\\ {\vspace{-1ex}\footnotesize\sffamily\mdseries \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf} {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}

### 2007-10-26 Developer/ProgrammingGuide/paper.tex
Update Programming Guide, Release Notes and add release annoucement.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{4cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
  \textbf{DRAFT}\\[\medskipamount]
  Version 0.9.3 for Equalizer 0.4\\[\medskipamount]
  \today
}

@@

OpenGL is a registered trademark, OpenGL Multipipe is a trademark of
Silicon Graphics, Inc. Linux is a registered trademark of Linus
Torvalds. CAVELib is a registered trademark of the University of
Illinois. The CAVE is a registered trademark of the Board of Trustees of
the University of Illinois at Chicago. All other trademarks and
copyrights herein are the property of their respective owners.

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.9.3   & Oct 21, 2007 & update title page, add front matter\\
    0.9.2   & Oct 19, 2007 & Jonas' feedback, changes related to
                             per-node frame sync\\
    0.9.1   & Oct 8,  2007 & review pass\\
    0.9     & Oct 7,  2007 & add eqHello, draft assembly in eVolve\\
    0.8     & Sep 28, 2007 & add head tracking, finish channel,
                             proof-reading pass\\
    0.7     & Sep 21, 2007 & start channel and image compositing,
                             add event handling, move to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & add pipe, window and object manager\\
    0.5     & Sep 3,  2007 & add config and partly pipe\\
    0.4     & Aug 31, 2007 & add distributed objects\\
    0.3     & Aug 26, 2007 & add application and render client\\
    0.2     & Aug 20, 2007 & add main function\\
    0.1     & Aug 19, 2007 & outline the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer provides a framework for the development of parallel OpenGL
applications. Equalizer-based applications can run a single
shared-memory system with multiple graphics cards (GPU's) or on graphics
clusters. This Programming Guide introduces the programming interface
using the \textsf{eqPly} example shipped with Equalizer as a guideline.

Equalizer is the next step in the evolution of generic parallel programming
interfaces for visualization applications using OpenGL. Existing
% rephrase ( aber ich weiss auch nicht wie :) )
solutions, such as OpenGL Multipipe SDK, Cavelib and VRJuggler,
implement a subset of concepts similar to Equalizer. In other areas,
e.g., tracking device support, they provide more functionality.

Equalizer implements the minimum necessary layer to build parallel and
scalable OpenGL applications. The programmer structures the application
so that the OpenGL rendering can be executed in parallel, potentially
using multiple processes for cluster-based execution. Equalizer provides
the domain-specific parallel rendering know-how and abstracts
configuration, threading, synchronization, windowing and event
handling. It is a `GLUT on steroids', providing parallel and distributed
execution, scalable rendering features and fully customizable event
handling.

If you have any question regarding Equalizer programming, this programming
guide, or specific problems you encountered, please direct them to the 
 \textsf{eq-dev} mailing list\footnote{see \link{http://www.equalizergraphics.com/lists.html}}.

@@

Equalizer can be installed by downloading the
distribution\footnote{\link{http://www.equalizergraphics.com/downloads.html}}
and by compiling the source code. After installing Equalizer, please
take a look at the Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
to get familiar with the capabilities of the \textsf{eqPly} example.

@@

The Equalizer architecture is based on a client-server model. Its client
library exposes all functionality discussed in this document to the
programmer, and provides communication between the different Equalizer
processes.

\subsubsection{Server}
Each Equalizer server is responsible for managing one visualization
system, i.e., a shared memory system or graphics cluster.  It controls
and launches the application's rendering clients.  Currently, Equalizer
currently only supports one application per server, but it will provide
concurrent and efficient multi-application support in future.

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an
application. Its execution is passive, it has no main loop and
is completely driven by rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{ssTaskMethods}) in the correct thread and context.

The application might contain a rendering client, in which case it can 
also contribute to the rendering. If it does not implement any render
client-related code, it is reduced to be the application's `master'
process without any OpenGL windows.

The rendering client can be the same executable as the application, as
it is the case with \textsf{eqPly}. When it is started as a render client,
the Equalizer initialization routine does not return and takes over the
control by calling the render client task methods. Complex real-world
applications often implement a separate, light-weight rendering client.

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/eqHello.png}
  {\caption{\small\label{fHello}Hello, World!}}
\end{wrapfigure}

The \textsf{eqHello} example is a minimal example to illustrate the
basic principle of an Equalizer application: The application developer
has to implement the rendering method \textsf{Channel::frameDraw},
similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
run as a stand-alone application from the command line.

@@

Equalizer uses a C++ programming interface. The API is minimally
invasive, so Equalizer imposes only a minimal, natural execution
framework upon the application. It does not provide a scene graph, or
interfere in any other way with the application's rendering code. The
restructuring work required for Equalizer is needed anyway in order to
parallelize the application for rendering.

@@

The application inherits from Equalizer classes and overrides virtual
functions to implement certain functionality, such as the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are similar in concept to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list can be found on the website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

\if 0
\subsection{Thread Safety}

Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains one current context for each
thread. It is difficult, if not impossible, to render from different
threads to the same context.

The rendering threads concurrently render the application's data
base. The data base should be accessed in a read-only fashion during
rendering, as it is the case with all modern scene graphs.

The rendering threads run asynchronously to the application's main
thread. Depending on the configuration's latency, the fall $n$ frames
behind the last frame finished by the application thread.

TBD support and explain per-node frame sync.
\fi

\subsection{\label{sResource}The Resource Tree}

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/cave.pdf}
  {\caption{\small\label{fConfig}An example configuration}}
\end{figure}

@@

\subsubsection{Configuration}

%FIXME confusing
The root of the resource tree is the \textsf{eq::Config}, which
represents the current configuration of the application. The
configuration is the session in which all render clients are
registered. The config instance on a given node currently only holds the
local node, not all nodes of the configuration.

@@

An \textsf{eq::Node} is the representation of a single computer in the
system. One operating system process of the render client will be used
for each node. Each configuration might also use an application node, in
which case the application process is also used for rendering. All
node-specific task methods are executed from the main application
thread.

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU).  In
the current implementation it is also one operating system
thread. Non-threaded pipes are supported for integrating with
thread-unsafe libraries, but have various performance caveats. All pipe,
window and channel task methods are executed from the pipe's thread or
in the case of non-threaded pipes, from the main application
thread\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

An \textsf{eq::Window} holds a drawable and an OpenGL context. The
drawable can be an on-screen window or an off-screen\footnote{Off-screen drawables are not implemented yet, but can be
  created by the application and used with Equalizer} PBuffer or
FBO(Framebuffer object). The window holds
window-system-specific handles to the drawable and context, e.g. an X11
window \textsf{XID} and \textsf{GLXContext} for the glX window system.

@@

\subsection{Compounds}

Usage of the rendering resources is configured using a compound 
tree. Although the API does not currently expose compounds, the
basic design behind compounds is explained here for a better 
understanding of the Equalizer architecture.

Each compound has a channel, which it uses to execute the rendering
tasks. One channel might be used by multiple compounds. Unused channels
are not instantiated. The rendering tasks for the channels are computed
by the server and send to the appropriate render clients.

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/frustra.pdf}
  {\caption{\small\label{fFrustra}Frustrum descriptions}}
\end{wrapfigure}
The channels of the leaf compounds in the compound tree are designated
as source channels, and the channel of the top-most compound is the
destination channel. Source channels execute the rendering for the
destination channel. The top-most compound with a channel defines the 2D
pixel viewport rendered by all leaf compounds. The destination channel
and pixel viewport can not be overridden.

Compounds have a frustum description, which is inherited to the
children. The frustrum can be specified as a wall, which is completely
defined by the bottom-left, bottom-right and top-left coordinates
relative to the origin, or as a projection, which is defined by the
position and head-pitch-roll orientation of the projector, as well as
the horizontal and vertical field-of-view and distance of the projection
wall. \fig{fFrustra} illustrates the wall and projection frustum
parameters.

Compounds have attributes which configure the decomposition of the
destination's channel viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in
screen space. A \textsf{range} tells the application to render a part of
its database, and an eye pass can selectively render different stereo
passes.

Compounds have output and input frames which configure the recomposition
of the resulting pixels from the source compound channels. An output
frame connects to an input frame of the same name, which transports the
selected frame buffer data from the output channel to the input
channel. This composition process is extensively described in
\sref{sCompositing}.

More information on the configuration of compounds can be found on the
Equalizer website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.

@@

The \textsf{eqPly} example is shipped with the Equalizer distribution
and serves as a reference implementation of an Equalizer-based
application of medium complexity. Its focus lies not on rendering features
or visual quality, but on providing a working show case for the integration
of most of the Equalizer features.

In this section the source code of \textsf{eqPly} is discussed in
detail, and relevant design decision, caveat and remarks are raised.

All classes in the example are in the \textsf{eqPly} namespace to avoid
type name ambiguities, in particular for the \textsf{Window} class which
is frequently used as a type in the global namespace by windowing
systems. \fig{fUml} shows the most important Equalizer classes and how
they are subclassed by the \textsf{eqPly} example.

The derived classes fall into two categories: The subclassed rendering
entities as discussed in \sref{sResource}, and two classes for
distributing data. For each of the rendering entities, its function and
typical usage will be discussed. The distributed data classes illustrate
the typical usage of static and dynamic, frame-specific data.

@@

@@

The main function starts off with parsing the command line into the
\textsf{LocalInitData} data structure, part of which will be distributed
to all render client nodes. The actual command line parsing is done by
the \textsf{LocalInitData} class and will be discussed in
\sref{sInitData}:

@@

The node factory is used by Equalizer to create the objects or instances for
the rendering entities. Each of the classes inherits from the same type
provided by Equalizer in the \textsf{eq} namespace. The provided
\textsf{eq::NodeFactory} base class instantiates a 'plain' Equalizer
object, thus making it possible to selectively subclass individual
entity types, as it is done by \textsf{eqHello}. For each rendering
resource used in the configuration, one C++ object will be created:

@@

The third step is to create an instance of the application and to
initialize it locally. The application is an \textsf{eq::Client}, which
in turn is an \textsf{eqNet::Node}. The underlying network distribution
in Equalizer is a peer-to-peer network structure of
\textsf{eqNet::Node}s. The application programmer does not have to be
aware of the classes in the \textsf{eqNet} namespace, but both the
\textsf{eq::Client} and the server are \textsf{eqNet::Node}s. The local
initialization of nodes creates a local listening socket, so that the
node, and therefore the \textsf{eq::Client} can communicate over the
network with other nodes, such as the server and the rendering clients.

@@

Finally everything is set up to run the \textsf{eqPly} application:

@@

After finishing, the application and \textsf{eqPly} are
de-initialized and the \textsf{main} function returns:

@@

\begin{enumerate}
  \item \textbf{Application}: The executable started by the user, which
    is the controlling entity in the rendering session.
  \item \textbf{Auto-launched render client}: The typical render client,
    started by the server. The server starts the executable with special
    parameters, which cause \textsf{Client::initLocal} to never
    return. During exit, the server terminates the process. By default,
    the server starts render client using \textsf{ssh}.
  \item \textbf{Resident render client}: Manually pre-started render
    client, listening on a specified port for server commands. This mode
    is selected using the command-line option \textsf{--eq-client} and
    potentially \textsf{--eq-listen} and \textsf{-r}\footnote{see \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

The application's main loop starts by connecting the application to an
Equalizer server. The command line parameter \textsf{--eq-server}
explicitly specifies a server address. If no server was specified,
\textsf{Client::connectServer} try first to connect to a server on the
local machine using the default port 4242. If that fails, it will create
a server running within the applications process with a default
1-channel configuration\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/standalone.html}}.

@@

The second step is to ask the server for a configuration. The
\textsf{ConfigParams} are a placeholder for later Equalizer
implementations to provide additional hints and information to the
server for choosing a configuration. The configuration chosen by the
server is created locally using
\textsf{NodeFactory::createConfig}. Therefore it is of type
\textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
making cast necessary:

@@

When the configuration was successfully initialized, the main rendering
loop is executed. The main loop runs until either the user exits the
configuration or a maximum number of frames has been rendered, specified
as a command-line argument. The latter is useful for benchmarks. The
\textsf{Clock} is reused for measuring the overall performance. A new
frame is started using \textsf{Config::startFrame} and a frame is
finished using \textsf{Config::finishFrame}.

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames. This allows
overlapping execution in the node and pipe threads and minimizes idle
times. The first \textsf{nLatency Config::finishFrame} return
immediately, since they have no frame to synchronize
upon. \fig{fSyncAsync} shows the execution of (hypothetical) rendering
tasks without latency~(\fig{fSync}) and with a latency of
one~(\fig{fAsync}).

With \textsf{eqPly}, a speedup of 15\% has been observed on a five-node
rendering
cluster when using a latency of one instead of no latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}.

@@

Equalizer provides distributed objects which help implementing data
distribution in a graphics cluster. The master version of a distributed
object is registered with a \textsf{eqNet::Session}, which assigns a
session-unique identifier to the object. Other nodes can map their
instance of the object to this identifier, thus synchronizing the
object's data with the remotely registered object.

@@

The \textsf{eqPly} example has a static distributed object to provide
initial data to all rendering nodes, as well as a versioned object to
provide frame-specific data such as camera position to the
rendering methods.

@@

On the application side, the class \textsf{LocalInitData} subclasses
\textsf{InitData} to provide the command line parsing and to set the
initial values. The render nodes only instantiate the distributed part
in \textsf{InitData}

A static distributed object either has to provide a pointer and size to
its data using \textsf{setInstanceData}, or it has to implement
\textsf{getInstanceData} and \textsf{applyInstanceData}. The first
approach can be used if all distributed member variables can be stored
in one contiguous block of memory. The second approach is used
otherwise.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 0.9.4 for Equalizer 0.4\\[\medskipamount]
  \today
}

@@

OpenGL is a registered trademark, OpenGL Multipipe is a trademark of
Silicon Graphics, Inc. Linux is a registered trademark of Linus
Torvalds.  Mac OS is a trademark of Apple Inc. CAVELib is a registered
trademark of the University of Illinois. The CAVE is a registered
trademark of the Board of Trustees of the University of Illinois at
Chicago. Qt is a registered trademark of Trolltech. All other
trademarks and copyrights herein are the property of their respective
owners.

@@

\vfill

\subsection*{Previous Page}

The images on the front page show: a terrain rendering application on a
six-node display wall [top left], the \textsf{eqPly} polygonal renderer
in a three-sided CAVE [middle], two volume rendering results from
eVolve\footnote{Data sets courtesy of General Electric, USA and AVS,
  USA} [bottom left and middle] and a six-node sort-last database
decomposition with parallel direct-send recomposition\footnote{Data set
  courtesy of Stanford University Computer Graphics Laboratory} [bottom
right].

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.9.4   & Oct 26, 2007 & release candidate 1\\
    0.9.3   & Oct 21, 2007 & update title page, add front matter\\
    0.9.2   & Oct 19, 2007 & Jonas' feedback, changes related to
                             per-node frame sync\\
    0.9.1   & Oct 8,  2007 & review pass\\
    0.9     & Oct 7,  2007 & add eqHello, draft assembly in eVolve\\
    0.8     & Sep 28, 2007 & add head tracking, finish channel,
                             proof-reading pass\\
    0.7     & Sep 21, 2007 & start channel and image compositing,
                             add event handling, move to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & add pipe, window and object manager\\
    0.5     & Sep 3,  2007 & add config and partly pipe\\
    0.4     & Aug 31, 2007 & add distributed objects\\
    0.3     & Aug 26, 2007 & add application and render client\\
    0.2     & Aug 20, 2007 & add main function\\
    0.1     & Aug 19, 2007 & outline the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer provides a framework for the development of parallel OpenGL
applications. Equalizer-based applications can run from a single
shared-memory system with a one or multiple graphics cards up to
large-scale graphics clusters. This Programming Guide introduces the
programming interface, often using the Equalizer \textsf{eqPly} example
as a guideline.

Equalizer is the next step in the evolution of generic parallel programming
interfaces for OpenGL-based visualization applications. Existing
solutions, such as OpenGL Multipipe SDK, Cavelib and VRJuggler,
implement a subset of concepts similar to Equalizer. In other areas,
e.g., tracking device support, they provide more functionality.

In order to adapt an application for Equalizer, the programmer
structures the source code so that the OpenGL rendering can be executed
in parallel, potentially using multiple processes for cluster-based
execution. Equalizer provides the domain-specific parallel rendering
know-how and abstracts configuration, threading, synchronization,
windowing and event handling. It is a `GLUT on steroids', providing
parallel and distributed execution, scalable rendering features and
fully customizable event handling.

If you have any question regarding Equalizer programming, this
programming guide, or other specific problems you encountered, please
direct them to the \textsf{eq-dev} mailing list\footnote{see
  \link{http://www.equalizergraphics.com/lists.html}}.

@@

Equalizer can be installed by downloading the
distribution\footnote{\link{http://www.equalizergraphics.com/downloads.html}}
and compiling the source code. After installing Equalizer, please take a
look at the Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
to get familiar with the capabilities of the \textsf{eqPly} example.

@@

The Equalizer architecture is based on a client-server model. The client
library exposes all functionality discussed in this document to the
programmer, and provides communication between the different Equalizer
processes.

\subsubsection{Server}
Each Equalizer server is responsible for managing one visualization
system, i.e., a shared memory system or graphics cluster.  It controls
and launches the application's rendering clients.  Currently, Equalizer
only supports one application per server, but it will provide
concurrent and efficient multi-application support in future.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an application. Its
execution is passive, it has no main loop and is completely driven by
Equalizer, based on the rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{ssTaskMethods}) in the correct thread and context. The application
either implements the task methods with application-specific code or
uses the default methods provided by Equalizer.

The application can also be a rendering client, in which case it can
also contribute to the rendering. If it does not implement any render
client-related code, it is reduced to be the application's `master'
process without any OpenGL windows and rendering code.

The rendering client can be the same executable as the application, as
it is the case with all provided examples. When it is started as a
render client, the Equalizer initialization routine does not return and
takes over the control by calling the render client task
methods. Complex applications usually implement a separate, light-weight
rendering client.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eqHello.png}
  {\caption{\small\label{fHello}Hello, World!}}
\end{wrapfigure}

The \textsf{eqHello} example is a minimal application to illustrate the
basic principle of an Equalizer application: The application developer
has to implement the rendering method \textsf{Channel::frameDraw},
similar to the \textsf{glutDisplayFunc} in GLUT applications. It can be
run as a stand-alone application from the command line.

@@

Equalizer uses a C++ programming interface. The API is minimally
invasive, so Equalizer imposes only a minimal, natural execution
framework upon the application. It does not provide a scene graph, or
interfere in any other way with the application's rendering code. The
restructuring work required for Equalizer is the minimal refactoring
needed to parallelize the application for rendering.

@@

The application inherits from Equalizer classes and overrides virtual
functions to implement certain functionality, e.g., the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are similar in concept to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list can be found on the website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

\subsection{Execution Modeal and Thread Safety}

Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchonization or protection of shared data.

The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{model}

The rendering threads concurrently render the application's
database. The data\-base should be accessed in a read-only fashion
during rendering to avoid threading problems. This is normally the case,
for example all modern scene graphs use read-only render traversals.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially.

Rendering threads on a single node are by default synchronized. When a
frame is finished, all local rendering threads are done
drawing. Therefore the application can safely modify the data between
the end of a frame and the beginning of a new frame. Furthermore, only
one instance of the application data has to be maintained within a
process since all rendering threads are quaranteed to draw the same frame.

This per-node frame synchronization does not inhibit latency across
rendering nodes. Furthermore, advanced rendering software which
multi-buffers the dynamic parts of the database can disable the per-node
frame synchronization, as explained in \sref{sNodeFrame}. Some scene
graphs do implement multi-buffered data.

The \textsf{eq::Config} represents the current configuration of the
application. The configuration is the session in which all render
clients are registered. A configuration consists of the description of
the rendering resources and the usage description for these resources.

@@

The resource usage is configured using a compound tree, which is a
hierarchical representation of the rendering decomposition and
recomposition across the resources. It is explained in \sref{compounds}.

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/cave.pdf}
  {\caption{\small\label{fConfig}An example configuration}}
\end{figure}

@@

@@

An \textsf{eq::Node} is the representation of a single computer in a
cluster. One operating system process of the render client will be used
for each node. Each configuration might also use an application node, in
which case the application process is also used for rendering. All
node-specific task methods are executed from the main application
thread.

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU).  In
the current implementation it is also one operating system
thread. Non-threaded pipes are supported for integrating with
thread-unsafe libraries, but have various performance caveats. All pipe,
window and channel task methods are executed from the pipe's thread, or
in the case of non-threaded pipes from the main application
thread\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

An \textsf{eq::Window} holds a drawable and an OpenGL context. The
drawable can be an on-screen window or an off-screen\footnote{Off-screen drawables are not implemented yet, but can be
  created by the application and used with Equalizer} PBuffer or
framebuffer object (FBO). The window holds
window-system-specific handles to the drawable and context, e.g. an X11
window \textsf{XID} and \textsf{GLXContext} for the glX window system.

@@

\subsection{\label{compounds}Compounds}

Usage of the rendering resources is configured using a compound
tree. Although the API does not currently expose compounds, the basic
design behind compounds is explained here for a better understanding of
the Equalizer architecture. Further information on the configuration of
compounds can be found on the Equalizer website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/compounds.html}}.

\subsubsection{Compound Channels}
Each compound has a channel, which it uses to execute the rendering
tasks. One channel might be used by multiple compounds. Unused channels
are not instantiated during initialization. The rendering tasks for the
channels are computed by the server and send to the appropriate render
clients.

\subsubsection{Frustum}
\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frustra.pdf}
  {\caption{\small\label{fFrustra}Wall and projection parameters}}
\end{wrapfigure}
Compounds have a frustum description to define the physical layout of
the display environment.

The frustum description is inherited by the children, therefore the
frustum is typically defined on the top-most compound.

The frustrum can be specified as a wall or projection description.

A wall is completely defined by the bottom-left, bottom-right and
top-left coordinates relative to the origin.

The projection is defined by the position and head-pitch-roll
orientation of the projector, as well as the horizontal and vertical
field-of-view and distance of the projection wall.

\fig{fFrustra} illustrates the wall and projection frustum parameters.

\subsubsection{Compound Classification}
The channels of the leaf compounds in the compound tree are designated
as source channels. The top-most channel in the tree is the destination
channel. Only source channels execute rendering task. All channels in a
compound tree work for the destination channel. The destination channel
defines the 2D pixel viewport rendered by all leaf compounds. The
destination channel and pixel viewport can not be overridden by child
compounds.

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination's channel viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in
screen space. A \textsf{range} tells the application to render a part of
its database, and an \textsf{eye} pass can selectively render different
stereo passes. Setting one or multiple attributes causes the parent's
view to be decomposed accordingly. Attributes are cumulative, that is,
intermediate compound attributes affect and therefore decompose the
rendering of all their children.

\subsubsection{Recomposition - Frames}
Compounds use output and input frames to configure the recomposition of
the resulting pixel data from the source channels. An output frame
connects to an input frame of the same name. The selected frame buffer
data is transported from the output channel to the input channel. The
assembly routine of the input channel will block on the availability of
the output frame. This composition process is extensively described in
\sref{sCompositing}.

@@

In this section the source code of \textsf{eqPly} is discussed in
detail, and relevant design decision, caveats and other remarks are
raised.

The \textsf{eqPly} example is shipped with the Equalizer distribution
and serves as a reference implementation of an Equalizer-based
application of medium complexity. It focuses on the example usage of
core Equalizer features, not on rendering features or visual quality.

@@

All classes in the example are in the \textsf{eqPly} namespace to avoid
type name ambiguities, in particular for the \textsf{Window} class which
is frequently used as a type in the global namespace by windowing
systems. \fig{fUml} shows how the most important Equalizer classes are
used through inheritance by the \textsf{eqPly} example.

The \textsf{eqPly} classes fall into two categories: Subclasses of the
rendering entities introduced in \sref{sResource}, and two other classes
for distributing data. The function and typical usage for each of the
rendering entities is discussed in this section.

The distributed data classes are helper classes based on
\textsf{eqNet::Object}. They illustrate the typical usage of distributed
objects for static as well as dynamic, frame-specific data.

@@

The main function starts off with parsing the command line into the
\textsf{LocalInitData} data structure. A part of it, the base class
\textsf{InitData}, will be distributed to all render client nodes. The
command line parsing is done by the \textsf{LocalInitData} class, which
is discussed in \sref{sInitData}:

@@

The node factory is used by Equalizer to create the object instances of
the configured rendering entities. Each of the classes inherits from the
same type provided by Equalizer in the \textsf{eq} namespace. The
provided \textsf{eq::NodeFactory} base class instantiates 'plain'
Equalizer objects, thus making it possible to selectively subclass
individual entity types, as it is done by \textsf{eqHello}. For each
rendering resources used in the configuration, one C++ object will be
created during initialization:

@@

The third step is to create an instance of the application and to
initialize it locally. The application is an \textsf{eq::Client}, which
in turn is an \textsf{eqNet::Node}. The underlying network layer in
Equalizer is a peer-to-peer network of \textsf{eqNet::Node}s. The
application programmer does not usually have to be aware of the classes
in the \textsf{eqNet} namespace, but both the \textsf{eq::Client} and
the server are \textsf{eqNet::Node}s.

The local initialization of a node creates a local listening socket, so
that the node. Therefore the \textsf{eq::Client} can communicate over
the network with other nodes, such as the server and the rendering
clients.

@@

Finally everything is set up, and the \textsf{eqPly} application is executed:

@@

After the application has finished, it is de-initialized and the
\textsf{main} function returns:

@@

\begin{enumerate}
\item \textbf{Application}: The executable started by the user, the
  controlling entity in the rendering session.
\item \textbf{Auto-launched render client}: The typical render client,
  started by the server. The server starts the executable with special
  parameters, which cause \textsf{Client::initLocal} to never
  return. During exit, the server terminates the process. By default,
  the server starts render client using \textsf{ssh}.
\item \textbf{Resident render client}: Manually pre-started render
  client, listening on a specified port for server commands. This mode
  is selected using the command-line option \textsf{--eq-client} and
  potentially \textsf{--eq-listen \textless address\textgreater} and
  \textsf{-r}\footnote{see
    \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

The application's main loop starts by connecting the application to an
Equalizer server. The command line parameter \textsf{--eq-server}
explicitly specifies a server address. If no server was specified,
\textsf{Client::connectServer} tries first to connect to a server on the
local machine using the default port. If that fails, it will create
a server running within the application process with a default
one-channel configuration\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/standalone.html}}.

@@

The second step is to ask the server for a configuration. The
\textsf{ConfigParams} are a placeholder for later Equalizer
implementations to provide additional hints and information to the
server for choosing the configuration. The configuration chosen by the
server is created locally using
\textsf{NodeFactory::createConfig}. Therefore it is of type
\textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
making the cast necessary:

@@

When the configuration was successfully initialized, the main rendering
loop is executed. The main loop runs until the user exits the
configuration, or when a maximum number of frames has been rendered,
when specified by a command-line argument. The latter is useful for
benchmarks. The \textsf{Clock} is reused for measuring the overall
performance. A new frame is started using \textsf{Config::startFrame}
and a frame is finished using \textsf{Config::finishFrame}.

@@

\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames. This allows
overlapping execution in the node and pipe threads and minimizes idle
times.

By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of exising rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization can disable it, as
explained in \sref{sNodeFrame}.

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
without latency (\fig{fSync}) and with a latency of
one (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
observed on a five-node rendering cluster when using a latency of one
instead of no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

Equalizer provides distributed objects which help implementing data
distribution in a cluster environment. The master version of a distributed
object is registered with a \textsf{eqNet::Session}, which assigns a
session-unique identifier to the object. Other nodes can map their
instance of the object to this identifier, thus synchronizing the
object's data with the remotely registered master version.

@@

The \textsf{eqPly} example has a static distributed object to provide
initial data to all rendering nodes, as well as a versioned object to
provide frame-specific data such as the camera position to the
rendering methods.

@@

On the application side, the class \textsf{LocalInitData} subclasses
\textsf{InitData} to provide the command line parsing and to set the
default values. The render nodes only instantiate the distributed part
in \textsf{InitData}.

A static distributed object either has to provide a pointer and size to
its data using \textsf{setInstanceData}, or it has to implement
\textsf{getInstanceData} and \textsf{applyInstanceData}. The first
approach can be used if all distributed member variables are stored in
one contiguous block of memory. The second approach is used otherwise.

The \textsf{InitData} class contains a string of variable
length. Therefore it uses the second approach of manually serializing
and deserializing its data. Serialization in \textsf{getInstanceData}
and deserialization in \textsf{applyInstanceData} is performed by
streaming all member variable to or from the provided data
streams. Efficient buffering and data transport between nodes is
implemented in the data streams:

{\footnotesize\begin{lstlisting}
void InitData::getInstanceData( eqNet::DataOStream& os )
{
    os << _frameDataID << _windowSystem << _useVBOs << _useGLSL << _filename;
}
